---
title: "Categorical Outcome"
---


```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
here::i_am("contents/categorical.qmd")
library(here)
```

```{r message=FALSE}
library(pwr)
library(ggplot2)
theme_set(theme_minimal())
```


```{r}
res.p.1 <- pwr.p.test(
  h = ES.h(p1 = 0.75, p2 = 0.50),
  sig.level = 0.05,
  power = 0.80,
  alternative = "greater"
)

res.p.1
```

```{r}
plot(res.p.1)
```

## Chi-square test

**Effect size:** Cohen’s W

A market researcher is seeking to determine preference among 4 package designs. He arranges to have a panel of 100 consumers rate their favorite package design. He wants to perform a chi-square goodness of fit test against the null of equal preference (25% for each design) with a significance level of 0.05. What's the power of the test if 3/8 of the population actually prefers one of the designs and the remaining 5/8 are split over the other 3 designs?

```{r}
null <- rep(0.25, 4)
alt <- c(3/8, rep((5/8)/3, 3))

ES.w1(null,alt)
```

### DF

$$
df = (l−1) × (k−1)
$$

Two categorical variables have $l$ categories and $k$ categories

### Power


```{r}
pwr.chisq.test(w=ES.w1(null,alt), N=100, df=(4-1), sig.level=0.05)
```



