---
title: "Population Proportion"
---



```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
here::i_am("contents/desc-proportion.qmd")
library(here)
```


## Concept

Cochran's sample size calculation is a **design tool** for determining the appropriate sample size to estimate a population proportion. As such, it does not directly test a hypothesis like a statistical test (e.g., t-test or ANOVA). However, the underlying **null and alternative hypotheses** behind Cochran’s sample size calculation are implicitly related to ensuring a certain **level of precision** (margin of error) for estimating proportions.

Below are the conceptual **null and alternative hypotheses** for both **infinite population** and **finite population adjustments**.


**1. Hypotheses for Infinite Population Proportion Sample Size Calculation**

- **Goal:**  Determine the minimum sample size needed to estimate a proportion ($p$) within a specified **margin of error (e)** and **confidence level (Z)** for a large population (assumed infinite).

- **Null Hypothesis ($H_0$)**: The estimated proportion ($\hat{p}$) **is not** within the desired margin of error ($e$) of the true population proportion ($p$) with the specified confidence level.

- **Alternative Hypothesis ($H_1$)**: The estimated proportion ($\hat{p}$) **is** within the desired margin of error ($e$) of the true population proportion ($p$) with the specified confidence level.


**2. Hypotheses for Finite Population Adjustment**

When calculating the sample size for a **finite population** size ($N$), we adjust the required sample size to account for the fact that the population is not infinitely large. This adjustment ensures efficiency (i.e., fewer participants required) without sacrificing precision.

- **Null Hypothesis ($H_0$)**: The estimated proportion ($\hat{p}$) **is not** within the desired margin of error ($e$) when adjusting for finite population size ($N$).

- **Alternative Hypothesis ($H_1$)**: The estimated proportion ($\hat{p}$) **is** within the desired margin of error ($e$) even with the adjustment for finite population size ($N$).


## Cochran’s Sample Size for Infinite Proportion

To estimate the sample size for an **infinite population**, we use **Cochran’s sample size formula for proportions**:


$$
n_0 = \frac{Z^2 \cdot p \cdot (1 - p)}{e^2}
$$


**Where:**

- $n_0$: Initial sample size (assuming an infinite population)
- $Z$: Z-value
- $p$: Expected proportion of the population with the attribute (use 0.5 for maximum variability)
- $e$: Margin of error ($e$ = 0.05)


The following R function implements Cochran’s formula:

```{r cochran_sample_size}
cochran_sample_size <- function(p, e, conf_level = 0.95) {
   
  Z <- qnorm(1 - (1 - conf_level) / 2) # Z-value for confidence level
  n0 <- (Z^2 * p * (1 - p)) / (e^2)    # Cochran's formula
  ceiling(n0)
  
}
```


### Example

Let’s say you want to calculate the required sample size to estimate a population proportion with: 

- **95% confidence level**
- **Estimated proportion (p) = 0.5** (maximum variability)
- **Margin of error (e) = 0.05**

```{r}
n0 <- cochran_sample_size(
  p = 0.5,           # Expected proportion (for maximum variability)
  e = 0.05,          # Margin of error
  conf_level = 0.95  # 95% confidence level
  )

n0
```

## Finite Population Correction

Since our study population is **finite**, we adjust the initial sample size using the **finite population correction formula**:

$$
n = \frac{n_0}{1 + \frac{n_0 - 1}{N}}
$$

**Where:**

- $n$: Adjusted sample size
- $N$: Total population size (total number of residents)


R function to apply this correction:


```{r finite_pop_adj}
finite_pop_adj <- function(n0, N) {
  n <- n0 / (1 + (n0 - 1) / N)
  ceiling(n) 
}
```

### Example

For a population of 1000, the required sample size is:

```{r}
finite_pop_adj(n0 = n0, N = 1000)
```

This approach helps you compute the sample size needed to accurately estimate a population proportion with a given confidence and margin of error.

